---
title: "Density and diffusion length data"
author: "Thomas Münch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density and diffusion length data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(FirnR)
library(PaleoSpec)
```

## Abstract

In this short vignette, we introduce the firn density data that is provided with
`FirnR` and showcase how to use package functions to obtain modelled density and
diffusion length data.

## Density data provided by FirnR

`FirnR` comes with density data of the firn cores B41 and B41 (Laepple et
al. 2016), which were drilled to a total depth of 112 m in close vicinity of
Kohnen Station in Dronning Maud Land, East Antarctica, during the field season
2012/2013 as part of the ANT-Land-2012 campaign of the COFI project conducted by
the Alfred Wegener Institute Helmholtz Centre for Polar and Marine Research. The
density data were obtained by means of high-resolution X-ray Computer Tomography
(CT; Freitag et al., 2013). Here, the data for the upper 25 m of the cores are
included.

The data set is provided in the variable `b41.b42.density` as a list of three
elements:

```{r}
names(b41.b42.density)
```

The elements `b41` and `b42` contain the density profiles of the firn cores B41
and B42, respectively, and the last element, `stack`, is the average density
data from both cores on the B41 depth scale, all provided as tibbles of two data
columns, `depth` in m and `density` in kg/m^3^:

```{r}
print(b41.b42.density$b41, n = 3)
print(b41.b42.density$b42, n = 3)
print(b41.b42.density$stack, n = 3)
```

Additionally, `stack` also contains a third column, `fitDensity`, which is a
second-order polynomial fit of the average density in the square root of the
depth.

To plot the data, we smooth it first with a 25 mm running mean filter.

```{r}
res <- 0.15
w   <- round(25 / res)
filter <- rep(1 / w, w)

b41.filtered <- b41.b42.density$b41
b41.filtered$density <- PaleoSpec::ApplyFilter(b41.filtered$density, filter,
                                               method = 2)

b42.filtered <- b41.b42.density$b42
b42.filtered$density <- PaleoSpec::ApplyFilter(b42.filtered$density, filter,
                                               method = 2)

stack.filtered <- b41.b42.density$stack
stack.filtered$density <- PaleoSpec::ApplyFilter(stack.filtered$density, filter,
                                                 method = 2)
stack.filtered$fitDensity <- PaleoSpec::ApplyFilter(stack.filtered$fitDensity,
                                                    filter, method = 2)

```

```{r, fig.width = 7, fig.height = 5.25}

plot(b41.filtered, type = "l", lwd = 1.5,
     xlab = "Depth (m)", ylab = "Density (kg/m3)")
lines(b42.filtered, col = "red", lwd = 1.5)

legend("bottomright", c("B41", "B42"), lty = 1, lwd = 1.5,
       col = c("black", "red"), bty = "n")
```

Plot the average data and compare it with the fit:

```{r, fig.width = 7, fig.height = 5.25}

plot(stack.filtered$depth, stack.filtered$density, type = "l", lwd = 1.5,
     xlab = "Depth (m)", ylab = "Density (kg/m3)")
lines(stack.filtered$depth, stack.filtered$fitDensity, col = "dodgerblue",
      lwd = 2.5)

legend("bottomright", c("Stack", "Fit"), lty = 1, lwd = 1.5,
       col = c("black", "dodgerblue"), bty = "n")
```

## Comparison with Herron-Langway densification

`FirnR` provides an implementation of the Herron-Langway densification model in
the function `DensityHL`. As input for the function, we need a depth vector on
which the density is modelled, and the site environmental conditions of
temperature [K], surface density [kg/m^3^] and accumulation rate
[kg/m^2^/yr]. For the latter, we use estimated environmental conditions for the
Kohnen Station drilling site.

```{r}
d <- seq(0, 25, 0.1)
density.hl <- data.frame(
  depth = d,
  density = DensityHL(d, rho.surface = 330, T = 273.15 - 40, bdot = 70)$rho
)
```

```{r, fig.width = 7, fig.height = 5.25}

plot(stack.filtered$depth, stack.filtered$density, type = "l", lwd = 1,
     xlab = "Depth (m)", ylab = "Density (kg/m3)", col = "lightgray")

lines(stack.filtered$depth, stack.filtered$fitDensity, col = "dodgerblue",
      lwd = 2.5)
lines(density.hl, col = "firebrick4", lwd = 2.5)

legend("bottomright", c("B41/B42 stack", "B41/B42 fit", "Herron-Langway"),
       lty = 1, lwd = c(1, 1.5, 1.5),
       col = c("lightgray", "dodgerblue", "firebrick4"), bty = "n")
```

## Calculation of firn diffusion lengths

With a density profile at hand, we can use `FirnR`'s function `DiffusionLength`
to model the diffusion length in firn, both for oxygen and hydrogen isotopes.

```{r}
sigma.d18O <- DiffusionLength(density.hl$depth, density.hl$density,
                              T = 273.15 - 40, P = 677, bdot = 70)
sigma.dD   <- DiffusionLength(density.hl$depth, density.hl$density,
                              T = 273.15 - 40, P = 677, bdot = 70,
                              dD = TRUE)
```

```{r, fig.width = 7, fig.height = 5.25}

plot(density.hl$depth, sigma.d18O, type = "n",
     xlab = "Depth (m)", ylab = "Diffusion length (cm)")

lines(density.hl$depth, sigma.d18O, col = "black", lwd = 1.5)
lines(density.hl$depth, sigma.dD, col = "dodgerblue4", lwd = 1.5)

legend("bottomright", c("d18O", "dD"), lty = 1, lwd = 1.5,
       col = c("black", "dodgerblue4"), bty = "n")
```

Since isotopic data from firn and ice cores are often dated, typically to annual
resolution, it is convenient to calculate the diffusion lengths directly in
units of time. This can be achieved with the function
`TemporalDiffusionLength`.

Note that the current approach here is to calculate the diffusion length in
units of length based on density from the Herron-Langway model, and to convert
this into diffusion length in time using the density and the given accumulation
rate; so it is not possible to directly calculate the temporal diffusion length
based on a given density profile.

The following is an example to obtain the diffusion length in years for a record
spanning the last 100 years. This choice is realised by setting the number of
time steps, `nt`, and the temporal resolution, `t.res`, accordingly.

```{r, fig.width = 7, fig.height = 5.25}
sigma.time <- TemporalDiffusionLength(nt = 100, t.res = 1,
                                      T = 273.15 - 40, P = 677, bdot = 70,
                                      rho.surface = 330)
```

```{r, fig.width = 7, fig.height = 5.25}
plot(sigma.time, type = "l", lwd = 1.5,
     xlab = "Time (yr)", ylab = "Diffusion length (yr)")

```

## References

Freitag, J., Kipfstuhl, S., and Laepple, T.: Core-scale radioscopic imaging:
a new method reveals density-calcium link in Antarctic firn, J. Glaciol., 59,
  1009–1014, https://doi.org/10.3189/2013JoG13J028, 2013

Laepple, T., Hörhold, M., Münch, T., Freitag, J., Wegner, A., and Kipfstuhl,
S.: Layering of surface snow and firn at Kohnen Station, Antarctica: Noise or
seasonal signal?, J. Geophys. Res. Earth Surf., 121, 1849–1860,
https://doi.org/10.1002/2016JF003919, 2016
