---
title: "Estimating and modelling temporal changes of firn properties and
profiles"
author: "Thomas Münch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating and modelling temporal changes of firn properties and profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(FirnR)
library(dplyr)
```

## Abstract

This vignette highlights the functions in `FirnR` which can be used to estimate
temporal changes, i.e. variations with depth, of firn properties, specifically
density and diffusion length, and to model how these changes affect firn proxy
records such as stable isotope profiles. In detail, this is studied in [Münch et
al. (2017)](https://doi.org/10.5194/tc-11-2175-2017), which also discusses the
exemplary estimates of densification rate, compression, and differential
diffusion length that are presented here.

## Densification rate

The function `EstimateDensificationRate` can be used to obtain an estimate of
the relative change in density over depth, obtained from a linear regression of
a given density profile.

The function returns the linear change in density relative to the surface
density, thus in units of an inverse length. In the following, we use the
average B41/B42 density profile provided with `FirnR` to estimate the
densification rate from the surface to a depth of 2 m, and to a depth of 5 m,
respectively, and express the results in % m^-1^:

```{R}
dr1 <- EstimateDensificationRate(b41.b42.density$stack, bottom.depth = 2)
print(round(dr1 * 1e2, 1))
dr2 <- EstimateDensificationRate(b41.b42.density$stack, bottom.depth = 5)
print(round(dr2 * 1e2, 1))
```

## Compression of firn profile

Under the assumption of a linear change in density with depth, thus given an
estimate of the relative densification rate, for instance from
`EstimateDensificationRate`, we can use the function `EstimateCompression` to
calculate the length change (i.e. compression) of a firn proxy record upon
advection downwards in the firn; as an example calculated in the following for a
1 m long profile advected downwards from the surface by an amount of 50 cm:

```{R}
EstimateCompression(top.depth = 0, length = 1, advection = 0.5, rate = dr1) %>%
  {{.} * 1e2} %>% # display in cm
  round(0)

EstimateCompression(top.depth = 0, length = 1, advection = 0.5, rate = dr2) %>%
  {{.} * 1e2} %>% # display in cm
  round(0)
```

## Change in diffusion length (differential diffusion length)

The differential diffusion length is a metric for the change in diffusion length
with depth, and it provides a concept for modelling the total diffusion a
stable isotope profile has undergone till some depth $z_2$ as the result of two
consecutive diffusional smoothings, the first during the advection from the
surface to a depth $z_1 < z_2$, and the second from $z_1$ to $z_2$ (see Münch et
al. (2017) for details).

Given a density profile and required environmental parameters, the differential
diffusion length can be obtained by first calculating a diffusion length profile
and then calling the function `GetDifferentialDiffusion`.

As an example, we calculate the diffusion length profile for
$\delta^{18}\mathrm{O}$ using the density data from the polynomial fit of the
average B41/B42 density profile, and calculate the differential diffusion length
(units of cm) over a time span of approximately 2 years given by the expected
change in top and bottom depths of the Kohnen trench profile T13 (cf. Münch et
al., 2017):


```{r}
d <- b41.b42.density$stack$depth
r <- b41.b42.density$stack$fitDensity
s <- DiffusionLength(d, r, P = 650, T = 228.5, bdot = 64)

data.frame(depth = d, sigma = s) %>%
  GetDifferentialDiffusion(z00 = 0, z01 = 1, z10 = 0.5, z11 = 1.5) %>%
  round(1)
```

## Modelling of firn profile changes

The main processes that affect firn profiles in the upper firn column are
advection, compression due to the increase of the density with depth, and, in
case of isotope profiles, diffusion (Münch et al., 2017). In `FirnR`, the
effects of these processes can be modelled with the functions `AdvectRecord`,
`CompressRecord`, and `DiffuseRecord`.

### Advection

The advection of a firn profile, i.e. the effective downward movement due to
fresh snowfall, is modelled by `AdvectRecord` as a simple shift in depth:

```{r, fig.width = 7, fig.height = 5.25}
# create some example firn profile
set.seed(20231211)
z <- 0 : 200
y <- -45 + 5 * sin(2 * pi * z / 50) + rnorm(length(z), 0.0001)

profile <- data.frame(depth = z, y = y)

plot(profile, xlab = "Depth (cm)", type = "l",
     ylab = "Some profile", xlim = c(0, 300))

profile %>%
  AdvectRecord(advection = 70, clip = FALSE) %>%
  lines(col = "dodgerblue", lwd = 1.5)

legend("topright", c("original", "advected"), col = c("black", "dodgerblue"),
       lty = 1, lwd = c(1, 1.5), bty = "n")
```

### Compression

Given an estimated length change of a firn profile due to densification, the
function `CompresseRecord` returns the compressed version of the profile
obtained from linear interpolation:

```{r, fig.width = 7, fig.height = 5.25}
plot(profile, xlab = "Depth (cm)", type = "l",
     ylab = "Some profile", xlim = c(0, 225))

profile %>%
  CompressRecord(compression = 30) %>%
  lines(col = "darkgreen", lwd = 1.5)

legend("topright", c("original", "compressed"), col = c("black", "darkgreen"),
       lty = 1, lwd = c(1, 1.5), bty = "n")
```

### Diffusion

The diffusion of stable isotope profiles can be modelled with the function
`DiffuseRecord`. The function expects a diffusion length either as a single
value, i.e. diffusion with a constant diffusion length, or as a vector giving
the change of the diffusion length over the depth of the isotope profile.

```{r, fig.width = 7, fig.height = 5.25}
sigma.const <- 8
sigma.depth <- seq(0, 4, length.out = length(z))

plot(profile, xlab = "Depth (cm)", type = "l", lwd = 2.5,
     ylab = "Some profile", xlim = c(0, 225))

profile.diff.const <- data.frame(
  depth = profile$depth,
  y     = DiffuseRecord(rec = profile$y, sigma = sigma.const, res = diff(z)[1])
)

lines(profile.diff.const, col = "darkorange3", lwd = 2.5)

profile.diff.depth <- data.frame(
  depth = profile$depth,
  y     = DiffuseRecord(rec = profile$y, sigma = sigma.depth, res = diff(z)[1])
)

lines(profile.diff.depth, col = "firebrick3", lwd = 2)

legend("topright", c("original", "const. diffusion", "incr. diffusion"),
       col = c("black", "darkorange3", "firebrick3"), seg.len = 1,
       lty = 1, lwd = c(2.5, 2.5, 2), bty = "n")
```

### Combination of processes

Finally, the function `ModifyRecord` provides a convenient way of changing a
profile according to the combination of all three modification processes, or
to some combination of them.

```{r, fig.width = 7, fig.height = 5.25}
plot(profile, xlab = "Depth (cm)", type = "l",
     ylab = "Some profile", xlim = c(0, 400))

profile %>%
  ModifyRecord(sigma = 3, compression = 50, advection = 200, clip = FALSE) %>%
  lines(col = "firebrick4", lwd = 1.5)

legend("topright", "original", col = "black", lwd = 1, bty = "n")
legend("bottomright", "advected + compressed + diffused",
       col = "firebrick4", lwd = 1.5, bty = "n")
```

## References

Münch, T., et al., Constraints on post-depositional isotope modifications
in East Antarctic firn from analysing temporal changes of isotope profiles,
The Cryosphere, 11(5), 2175-2188, https://doi.org/10.5194/tc-11-2175-2017,
2017.

